---
title: "ASSIGNMENT_5_MDTS4214_0731"
author: "Simlin saha"
date: "2026-02-26"
output: word_document
---

####3. Illustration of the Role of Ordinal Predictors in Multiple Linear Regression

###Consider the diamonds dataset available in the ggplot2 package in R.

###First, identify all ordinal (ordered categorical) variables in the dataset and determine the response variable.

###The ordered categorical predictors are:

###cut: Fair < Good < Very Good < Premium < Ideal

###color: D < E < F < G < H < I < J

###clarity: I1 < SI2 < SI1 < VS2 < VS1 < VVS2 < VVS1 < IF

###The response variable is price.

###(a)

###Fit a linear regression model with the response variable regressed on the quality of cut

```{r}
library(ggplot2)
library(MASS)
contrasts(diamonds$cut)=contr.sdif(5)

m1=lm(price ~ cut, data = diamonds)
summary(m1)
coef(m1)
```
####price=4062.24в€’429.89(Cut2в€’1)+52.89(Cut3в€’2)+602.50(Cut4в€’3)в€’1126.715(Cut5в€’4)
###(b) Test whether the mean price of diamonds with Premium cut differs significantly from that of Ideal cut.
###Hypotheses:рќђ»0:рќњ‡(Premium)=рќњ‡(Ideal)H0:Ој(Premium)=Ој(Ideal)
###versusрќђ»1:рќњ‡(Premium)в‰ рќњ‡(Ideal)H1:Ој(Premium)=Ој(Ideal)
###From the regression results, we observe:рќ‘ќ<2Г—10в€’16p<2Г—10в€’16 Since рќ‘ќ<0.05
###p<0.05, we reject рќђ»0

.
###Therefore, there is strong statistical evidence that the mean price of Premium cut diamonds is different from that of Ideal cut diamonds.
###Moreover,
###рќњ‡^(Ideal)в€’рќњ‡^(Premium)=в€’1126.72Ој^
###(Ideal)в€’Ој^	вЂ‹
###(Premium)=в€’1126.72
###This implies:
###рќњ‡^(Premium)в€’рќњ‡^(Ideal)=1126.72Ој^	вЂ‹(Premium)в€’Ој^

###(Ideal)=1126.72
###Hence, Premium cut diamonds are estimated to be approximately 1126.72 units more expensive on average compared to Ideal cut diamonds.
###(c) What is the expected price of a diamond with Ideal cut?

```{r}
predict(m1,
        newdata = data.frame(
          cut = factor("Ideal", levels=levels(diamonds$cut))
        ))
```
###### The expected price of a diamond with Ideal cut is approximately 3457.542

#### (d) Modify the regression model in (a) by incorporating the predictor вЂњtableвЂќ.Write the fitted regression model.

```{r}
m2=lm(price ~ cut + table, data = diamonds)
summary(m2)
coef(m2)
```

### price^ = в€’6340.256 в€’365.568(cut2-1) +185.163(cut3-2) +461.015(cut4-3) в€’626.220(cut5-4) +179.105(table)

##(e) Test for the significance of вЂњtableвЂќ in predicting the price of diamond.

### Hypotheses:

###рќђ»0:ОІ(table)=0
### against
### рќђ»1:ОІ(table)в‰ 0
### From the output, we have:
### рќ‘ќ<2Г—10^(в€’16)
### Since,рќ‘ќ<0.05,reject рќђ»0.
### So, table is highly significant for predicting diamond price (after accounting for cut).

## (f) Find the average estimated price of a diamond with an average table value and which is of fair cut.

```{r}
mean_table=mean(diamonds$table)
mean_table

predict(m2, newdata = data.frame(
  cut = factor("Fair", levels = levels(diamonds$cut)),
  table = mean_table
))
```
###So the average estimated price for a Fair cut diamond at table(mean) = 57.46 is: 4072.8

# K nearest neighbours regression
### Problem to demonstrate the utility of K nearest neighbour regression over least squares regression

## Consider a setting with n = 1000 observations. Generate
## (i) x1i from N(0, 2^2) and x2i from Poisson(О» = 1.5).
## (ii) Оµi from N(0, 1).
##(iii) yi = в€’2 + 1.4x1i в€’ 2.6x2i + Оµi.

## Split the data into train and test sets. Keep the first 800 observations as training data and the remaining as test data. Work out the following:

```{r}
set.seed(123)
n= 1000
x1=rnorm(n, mean=0, sd=2)
x2=rpois(n, lambda=1.5)
eps=rnorm(n, 0, 1)

y=-2 +1.4*x1 -2.6*x2 +eps

data=data.frame(x1, x2, y)

train=data[1:800, ]
test=data[801:1000, ]
```

## 1. Fit a multiple linear regression equation of y on x1 and x2. Calculate test MSE.

```{r}
lm_fit=lm(y ~ x1 + x2, data=train)
summary(lm_fit)

pred_lm=predict(lm_fit, newdata=test)

mse_lm=mean((test$y - pred_lm)^2)
mse_lm
```

### Test MSE в‰€ 1

## 2. Fit a KNN model with k = 1, 2, 5, 9, 15. Calculate test MSE for each choice of k.

```{r}
#install.packages("caret")
```

```{r}
library(caret)

k_values=c(1, 2, 5, 9, 15)
mse_knn_linear <- numeric(length(k_values))

for(i in seq_along(k_values)){
  knn_fit=knnreg(y ~ x1 + x2, data = train, k = k_values[i])
  y_pred_knn=predict(knn_fit, test)
  mse_knn_linear[i]=mean((test$y - y_pred_knn)^2)
}

data.frame(k = k_values, Test_MSE = mse_knn_linear)
```
### Conclusion:
### KNN with k = 1 may have a very low training MSE but can overfit, leading to higher test MSE.As k increases, the model becomes smoother, reducing variance but increasing bias.For small k (e.g., 1), test MSE may be higher due to high variance (overfitting).As k increases, MSE may decrease initially and then increase again.

### However, overall KNN does not outperform linear regression here.
###### Since the true relationship is linear, the parametric linear model is more efficient.

##Suppose the data in Step (iii) is generated as :
## yi = 1/(в€’2 + 1.4x1i в€’ 2.6x2i + 2.9x1i^2) + 3.1 sin(x2i) в€’1.5x1ix2i^2 + Оµi.

## Work out the problems in (1) and (2). Compare and comment on the results.

```{r}
y2=1/(-2 + 1.4*x1 - 2.6*x2 + 2.9*x1^2) +
      3.1*sin(x2) -
      1.5*x1*(x2^2) +
      eps

data2=data.frame(x1,x2,y=y2)

train2=data2[1:800,]
test2=data2[801:1000,]
```

```{r}
lm_fit2 <- lm(y ~ x1 + x2, data=train2)
summary(lm_fit2)

pred_lm2 <- predict(lm_fit2, newdata=test2)

mse_lm2 <- mean((test2$y - pred_lm2)^2)
mse_lm2
```
### Since it is a Misspecified Model, it has a large MSE.

```{r}
mse_knn_nonlinear=numeric(length(k_values))

for(i in seq_along(k_values)){
  knn_fit_nl=knnreg(y ~ x1 + x2, data = train2, k = k_values[i])
  y_pred_knn_nl=predict(knn_fit_nl, test2)
  mse_knn_nonlinear[i]=mean((test2$y - y_pred_knn_nl)^2)
}

data.frame(k = k_values, Test_MSE = mse_knn_nonlinear)
```

### conclusion:
### * In the linear model, the best KNN is at moderate k (9) and gives low test MSE (~1.21).

###* In the nonlinear model, the best KNN is at very small k (1), because the relationship is complex and needs high flexibility.

### This clearly demonstrates the biasвЂ“variance tradeoff,i.e;
### * If k is small then we get, low bias & high variance
### * If k is large then we get, higher bias & lower variance

